---
title: "spatstat package for spatial point patterns"
format: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```


```{r}
#| message: false
library(spatstat)
library(sf)
library(rnaturalearth)
```

# Careating spatial point patterns

Create the observation window

```{r}
win <- owin(xrange = c(0,1), yrange = c(0,2))
win; str(win)
```

```{r spatstat-package-1}
plot(win)
```

Simulate 100 randome points in the window

```{r}
x <- runif(100, 0, 1)
y <- runif(100, 0, 2)
X <- ppp(x = x, y = y, window = win)
X
```

```{r spatstat-package-2}
plot(X); axis(1); axis(2)
```

```{r}
Window(X)
```

Alternatively

```{r spatstat-package-3}
X2 <- runifpoint(100, win)
plot(X2); axis(1); axis(2)
```

Adding associated information (marks)

Assign a numeric value to each event

Two alternatives:

```{r spatstat-package-4}
marks(X) <- 1:npoints(X)
X <- X %mark% 1:npoints(X)
plot(X)
```

Test if points lie in a given window

```{r spatstat-package-5}
win <- owin() # a unit window
marks(X) <- inside.owin(X, w = win)
plot(X); axis(1); axis(2)
```

# Converting between ppp and sf objects

## ppp to sf

Here, we show how to use the st_as_sf() function from sf to transform the longleaf data from sptatstat which contains the locations and sizes of longleaf pine trees, from ppp to sf class.

```{r}
X <- longleaf
str(X)
```

```{r spatstat-package-6}
plot(X)
```


```{r}
d <- data.frame(x = X$x, y = X$y, m = marks(X)) %>% 
  st_as_sf(coords = c("x", "y"))
```

## sf to ppp

```{r spatstat-package-7}
X <- as.ppp(st_coordinates(d), st_bbox(d))
X <- X %mark% d$m
plot(X)
```

Here, we show an example on how to create a spatial point pattern with the boundary of Brazil as observation window.

```{r spatstat-package-8}
map <- ne_countries(type = "countries", country = "Brazil",
                    scale = "medium", returnclass = "sf") %>% 
  st_transform(crs = "EPSG:29172")
win <- as.owin(map)
X <- runifpoint(100, win = win)
plot(X)
```




