---
title: "04-terra.Rmd"
output: html_document
date: "2025-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Terra package

```{r}
library(terra)
library(ggplot2)
library(tidyterra)
```

> terra classes:
>
> -   `SpatRaster`
> -   `SpatVector`

## Raster data

> -   `rast()` to create and read rasters
> -   `writeRaster()` to write data

```{r}
pathraster = system.file("ex/elev.tif", package = "terra")
r = rast(pathraster)
plot(r)
```

```{r}
r <- rast(ncol = 10, nrow = 10, 
          xmin = -150, xmax = -80, 
          ymin = 20, ymax = 60)
r
```

> get and set values with `values()`

```{r}
values(r) <- 1:ncell(r)
values(r)
```

> Multilayer

```{r}
r2 <- r * r
s <- c(r, r2)
```

```{r}
plot(s[[2]])
```

```{r}
plot(min(s))
plot(r + r + 10)
plot(round(r))
plot(r == 1)
```

## Vector data

> -   `vect()` to read a shapefile
> -   `writeVector()` to write to a file

```{r}
pathshp <- system.file("ex/lux.shp", package = "terra")
v <- vect(pathshp)
```

> Create a `SpatVector`

```{r}
long <- c(-0.118092, 2.349014, -3.703339, 12.496366)
lat <- c(51.509865, 48.864716, 40.416729, 41.902782)
longlat <- cbind(long, lat)

crspoints <- "+proj=longlat +datum=WGS84"

d <- data.frame(
  place = c("London", "Paris", "Madrid", "Rome"), 
  value = c(200, 300, 400, 500)
)

pts <- vect(longlat, atts = d, crs = crspoints)

pts
plot(pts)
```

## Cropping, masking and aggregating

First, we use the `worldclim_country()` function of the **geodata** package ([Hijmans et al. 2023](https://www.paulamoraga.com/book-spatial/references.html#ref-R-geodata)) to download global temperature data from the [WorldClim](https://www.worldclim.org/data/index.html) database. Specifically, we download monthly average temperature in degree Celsius by specifying the country (`country = "Spain"`), the variable mean temperature (`var = "tavg"`), the resolution (`res = 10`), and the path where to download the data to as a temporary file (`path = tempdir()`). Figure [4.2](https://www.paulamoraga.com/book-spatial/the-terra-package-for-raster-and-vector-data.html#fig:terratemp) shows maps of the monthly average temperature in Spain.

```{r}
r <- geodata::worldclim_country(country = "Spain", var = "tavg", 
                                res = 10, path = tempdir())
plot(r)
```

```{r}
r <- mean(r)
plot(r)
```

> Download Spain but remove Canary Islands

```{r}
map <- rnaturalearth::ne_states("Spain", returnclass = "sf")
map <- map[-which(map$region == "Canary Is."), ]
ggplot(map) + geom_sf()
```

> Crop to the original to the spatial extent of the map

```{r}
sextent <- terra::ext(map)
r <- terra::crop(r, sextent)
plot(r)
```

> Mask values outside to NA

```{r}
r <- terra::mask(r, vect(map))
plot(r)
```

> Aggregate to lower resolution

```{r}
r <- terra::aggregate(r, fact = 20, fun = "mean", na.rm = T)
plot(r)
```

## Extracting values at points

Here, we provide an example of the use of [`extract()`](https://rdrr.io/pkg/terra/man/extract.html) using a raster representing the elevation of Luxembourg, and a vector file with the divisions of Luxembourg from files in the **terra** package.

```{r}
lrast <- rast(system.file("ex/elev.tif", package = "terra"))
lpolys <- vect(system.file("ex/lux.shp", package = "terra"))
```

```{r}
points <- crds(centroids(lpolys))
```

```{r}
plot(lrast)
plot(lpolys, add = T)
points(points)
```

```{r}
points <- as.data.frame(points)
values_at_points <- extract(lrast, points)
cbind(points, values_at_points)
```

## Extract and average (Polygons)

```{r}
head(extract(lrast, lpolys, na.rm = T))
```

```{r}
# Extracted raster cells and percentage of area
# covered within each polygon
head(extract(lrast, lpolys, na.rm = TRUE, weights = TRUE))
```

```{r}
lpolys$avg <- extract(lrast, lpolys, mean, na.rm = T)$elevation
```

The area-weighted average raster values by polygon are obtained with [`extract()`](https://rdrr.io/pkg/terra/man/extract.html) setting `weights = TRUE`.

```{r}
lpolys$weightedavg <- extract(lrast, lpolys, mean,
                              na.rm = T, weights = T)$elevation
```

```{r}
lpolys |> ggplot() +
  geom_spatvector(aes(fill = avg)) +
  scale_fill_terrain_c()

lpolys |> ggplot() +
  geom_spatvector(aes(fill = weightedavg)) +
  scale_fill_terrain_c()
```
