---
title: "05-maps"
output: html_document
date: "2025-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sf)
```

# Making Maps

The areal data we map correspond to sudden infant deaths in the counties of North Carolina, USA, in 1974 and 1979 which are in the **sf** package ([Pebesma 2022a](https://www.paulamoraga.com/book-spatial/references.html#ref-R-sf)).

```{r}
nameshp <- system.file("shape/nc.shp", package = "sf")
d <- st_read(nameshp, quiet = T)
d$vble <- d$SID74
d$vble2 <- d$SID79
```

## ggplot2

```{r}
library(ggplot2)
library(viridis)
ggplot(d) + geom_sf(aes(fill = vble)) +
  scale_fill_viridis() + theme_bw()
```

```{r}
png("plot.png")
ggplot(d) + geom_sf(aes(fill = vble)) +
  scale_fill_viridis() + theme_bw()
dev.off()
```

```{r}
ggplot(d) + geom_sf(aes(fill = vble)) +
  scale_fill_viridis() + theme_bw()
```

```{r}
library(plotly)
g <- ggplot(d) + geom_sf(aes(fill = vble))
plotly::ggplotly(g)
```

## leaflet

```{r}
st_crs(d)$epsg
```

```{r}
d <- st_transform(d, 4326)
```

-   create color palette mapped to vble
-   create map with `leaflet()` and `addTiles()` for background
-   use `addPolygons()` for counties
-   add legend

```{r}
library(leaflet)
pal <- colorNumeric(palette = "YlOrRd", domain = d$vble)
l <- leaflet(d) |> addTiles() |> 
  addPolygons(color = "white", fillColor = ~ pal(vble), 
              fillOpacity = 0.8) |> 
  addLegend(pal = pal, values = ~vble, opacity = 0.8)
l
```

```{r}
l |> addMiniMap()
```

```{r}
library(htmlwidgets)
saveWidget(widget = l, file = "map.html")
```

## mapview

```{r}
library(mapview)
mapview(d, zcol = "vble")
```

```{r}
library(RColorBrewer)
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
mapview(d, zcol = "vble", map.types = "CartoDB.DarkMatter", 
        col.regions = pal, layer.name = "SDI")
```

```{r}
map1 <- mapview(d, zcol = "vble")
leaflet::addMiniMap(map1@map)
```

## mapview - side by side

```{r}
library(leaflet.extras2)
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
```

> -   Create a common legend
> -   Create maps
> -   Display with pipe

```{r}
at <- seq(min(c(d$vble, d$vble2)), max(c(d$vble, d$vble2)), 
              length.out = 8)

m1 <- mapview(d, zcol = "vble", map.types = "CartoDB.Voyager", 
              col.retions = pal, at = at)
m2 <- mapview(d, zcol = "vble2", map.types = "CartoDB.Voyager", 
              col.retions = pal, at = at)
m1 | m2
```

## leafsync

```{r}
m <- leafsync::sync(m1, m2)
m
```

```{r}
htmltools::save_html(m, "m.html")
```

## tmap

```{r}
library(tmap)
tmap_mode("plot")
tm_shape(d) + tm_polygons("vble")
```

## Point data

```{r}
library(maps)
d <- world.cities
d <- d[which(d$country.etc == "South Africa"), ]
d <- st_as_sf(d, coords = c("long", "lat"))
st_crs(d) <- 4326
```

```{r}
d1 <- world.cities[which(world.cities$country.etc == "South Africa"), ] |> 
  st_as_sf(coords = c("long", "lat"))
```

```{r}
d$population <- d$pop
d$size <- sqrt(d$pop) / 100
```

```{r}
ggplot(d) +
  geom_sf(aes(col = population, size = size)) +
  scale_color_viridis()
```

```{r}
pal <- colorNumeric(palette = "viridis", domain = d$population)
leaflet(d) |> 
  addTiles() |> 
  addCircles(lng = st_coordinates(d)[, 1], 
             lat = st_coordinates(d)[, 2], 
             radius = ~sqrt(population)*10, 
             color = ~pal(population), popup = ~name) |> 
  addLegend(pal = pal, values = ~population, position = "bottomright")
```

```{r}
d$size = sqrt(d$population)
mapview(d, zcol = "population", cex = "size")
```

```{r}
tmap_mode("view")
tm_shape(d) + tm_dots("population", scale = sqrt(d$population)/500, 
                      palette = "viridis")
```

## Raster data

```{r}
library(terra)
r <- rast(system.file("ex/elev.tif", package = "terra"))
```

```{r}
d <- st_as_sf(as.data.frame(r, xy = T), coords = c("x", "y"))
st_crs(d) <- 4326
ggplot(d) + geom_sf() +
  geom_raster(data = as.data.frame(r, xy = T), 
              aes(x = x, y = y, fill = elevation))
```

To use the **leaflet** and **mapview** packages, we transform the data from class `terra` to `RasterLayer` with the [`raster::brick()`](https://rdrr.io/pkg/raster/man/brick.html) function. Figure [5.11](https://www.paulamoraga.com/book-spatial/making-maps-with-r.html#fig:mapsrasterleaflet) shows the map of raster data created with **leaflet**.

```{r}
library(raster)
rb <- raster::brick(r)

pal <- colorNumeric("YlOrRd", values(r), 
                    na.color = "transparent")
leaflet() |> 
  addTiles() |> 
  addRasterImage(rb, colors = pal, opacity = 0.8) |> 
  addLegend(pal = pal, values = values(r), title = "elevation")
```

```{r}
mapview(rb, layer = "elevation")
```

## Mobility flows

```{r}
library(flowmapblue)
locations <- data.frame(
id = c(1, 2, 3),
name = c("New York", "London", "Rio de Janeiro"),
lat = c(40.713543, 51.507425, -22.906241),
lon = c(-74.011219, -0.127738, -43.180244)
)
```

```{r}
flows <- data.frame(
  origin = c(1, 2, 3, 2, 1, 3),
  dest = c(2, 1, 1, 3, 3 , 2),
  count = c(42, 51, 50, 40, 22, 42)
  )
```

```{r}
flowmapblue(locations, flows, mapboxAccessToken = NULL,
            clustering = TRUE, darkMode = TRUE, animation = FALSE)
```
