---
title: "08 - Spatial Autocorrelation"
output: html_document
date: "2025-04-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Spatial Autocorrelation

```{r}
library(spdep)
library(spData)
library(sf)
library(mapview)
library(tmap)
```

Spatial autocorrelation can be assessed using indices that summarize the degree to which similar
observations tend to occur near each other over the study area. Two common indices that are used to assess spatial autocorrelation in areal data are Moran’s I ([Moran 1950](https://www.paulamoraga.com/book-spatial/references.html#ref-moran50)) and Geary’s C ([Geary 1954](https://www.paulamoraga.com/book-spatial/references.html#ref-geary54)).

In this chapter, we use the Moran’s I to test the spatial autocorrelation of housing prices in 506 census tracts in Boston, USA, in 1978 obtained from the **spData** package ([Bivand, Nowosad, and Lovelace 2022](https://www.paulamoraga.com/book-spatial/references.html#ref-R-spData)). The data has a variable called `MEDV` with the median prices of owner-occupied housing in \$1000 USD. We create the variable `vble` with the values of `MEDV` that will be used in the analysis. Figure [8.2](https://www.paulamoraga.com/book-spatial/spatial-autocorrelation.html#fig:arealsacdata) shows the map created with the housing prices using **mapview** ([Appelhans et al. 2022](https://www.paulamoraga.com/book-spatial/references.html#ref-R-mapview)).

```{r}
map <- st_read(system.file("shapes/boston_tracts.gpkg", 
                           package = "spData"))
map$vble <- map$MEDV
mapview(map, zcol = "vble")
```

## Global Moran's $I$

> Range from -1 (negative spatial autocorrelation) to +1 (postitive autocorrelation
>
> $I$ has a normal distribution with large number of regions

1.  State the null and alternative hypotheses:\
    H0:I=E[I] (no spatial autocorrelation),\
    1:I≠E[I] (spatial autocorrelation).
2.  Choose the significance level α we are willing to tolerate, which represents the maximum value for the probability of incorrectly rejecting the null hypothesis when it is true (usually α=0.05).
3.  Calculate the test statistic:\
    z=I−E(I)Var(I)1/2.
4.  Find the p-value for the observed data by comparing the z-score to the standard normal distribution or via Monte Carlo randomization.The p-value is the probability of obtaining a test statistic as extreme as or more extreme than the one observed test statistic in the direction of the alternative hypothesis, assuming the null hypothesis is true.
5.  Make one of these two decisions and state a conclusion:
    - If p-value <α, we reject the null hypothesis. We conclude data provide evidence for the alternative hypothesis.\
    - If p-value ≥α, we fail to reject the null hypothesis. The data do not provide evidence for the alternative hypothesis.

## `moran.test()`

> `alternative` set to `greater` (default), `less` or `two.sided`

```{r}
nb <- poly2nb(map, queen = TRUE)
nbw <- nb2listw(nb, style = "W")

gmoran <- moran.test(map$vble, nbw, 
                     alternative = "greater")
gmoran
```

```{r}
str(gmoran)
```

```{r}
gmoran[["estimate"]][["Moran I statistic"]]
```

> `statistic` is the z-score

```{r}
gmoran[["statistic"]]
```

```{r}
gmoran[["p.value"]]
```

## Monte Carlo simulation

```{r}
gmoranMC <- moran.mc(map$vble, nbw, nsim = 999)
gmoranMC
```

```{r}
str(gmoranMC)
```

```{r}
hist(gmoranMC$res)
```

## Moran's $I$ scatterplot

```{r}
moran.plot(map$vble, nbw)
```

## Local Moran's $I$

Local Indicators of Spatial Association (LISA) ([Anselin 1995](https://www.paulamoraga.com/book-spatial/references.html#ref-anselin95)) are designed to provide an indication of the extent of significant spatial clustering of similar values around each observation. A desirable property is that the sum of the LISA’s values across all regions is equal to a multiple of the global indicator of spatial association. As a result, global statistics may be decomposed into a set of local statistics and most LISAs are defined as local versions of well-known global indexes.

High values suggest an area is part of a cluster of similar observations.Low values indicate significant difference from neighbors.

## `localmoran()`

The [`localmoran()`](https://r-spatial.github.io/spdep/reference/localmoran.html) function of the **spdep** package can be used to compute the Local Moran’s I

for a given dataset.
The arguments of [`localmoran()`](https://r-spatial.github.io/spdep/reference/localmoran.html)
include a numeric vector with the values of the variable, a list with
the neighbor weights, and the name of an alternative hypothesis that can
be set equal to `greater` (default), `less` or `two.sided`.
The returned object of the [`localmoran()`](https://r-spatial.github.io/spdep/reference/localmoran.html) function contains the following information:

-   `Ii`: Local Moran’s I statistic for each area,
-   `E.Ii`: Expectation Local Moran’s Istatistic,
-   `Var.Ii`: Variance Local Moran’s I statistic,
-   `Z.Ii`: z-score,
-   `Pr(z > E(Ii))`, `Pr(z < E(Ii))` or `Pr(z != E(Ii))`: p-value for an alternative hypothesis `greater`, `less` or `two.sided`, respectively.

```{r}
lmoran <- localmoran(map$vble, nbw, alternative = "greater")
head(lmoran)
```

```{r}
map$lmI <- lmoran[, "Ii"] # local Moran's I
map$lmZ <- lmoran[, "Z.Ii"] # z-scores
map$lmp <- lmoran[, "Pr(z > E(Ii))"]
```

```{r, warning=FALSE}
p1 <- tm_shape(map) +
  tm_polygons(col = "vble", title = "vble", style = "quantile") +
  tm_layout(legend.outside = TRUE)

p2 <- tm_shape(map) +
  tm_polygons(col = "lmI", title = "Local Moran's I",
              style = "quantile") +
  tm_layout(legend.outside = TRUE)

p3 <- tm_shape(map) +
  tm_polygons(col = "lmZ", title = "Z-score",
              breaks = c(-Inf, 1.65, Inf)) +
  tm_layout(legend.outside = TRUE)

p4 <- tm_shape(map) +
  tm_polygons(col = "lmp", title = "p-value",
              breaks = c(-Inf, 0.05, Inf)) +
  tm_layout(legend.outside = TRUE)

tmap_arrange(p1, p2, p3, p4)
```

```{r}
tmap_mode("view")
tmap_arrange(p1, p2, p3, p4)
```

```{r}
tmap5 <- tm_shape(map) + 
  tm_polygons(col = "lmZ",
              title = "Local Moran's I", 
              style = "fixed",
              breaks = c(-Inf, -1.96, 1.96, Inf),
              labels = c("Negative SAC", "No SAC", "Positive SAC"),
              palette =  c("blue", "white", "red")) +
  tm_layout(legend.outside = TRUE)

ttm()
tmap5
```

## Clusters

Local Moran’s I allows us to identify clusters of the following types:

-   High-High: areas of high values with neighbors of high values,
-   High-Low: areas of high values with neighbors of low values,
-   Low-High: areas of low values with neighbors of high values,
-   Low-Low: areas of low values with neighbors of low values.

```{r}
lmoran <- localmoran(map$vble, nbw, alternative = "two.sided")
head(lmoran)
```

```{r}
map$lmp <- lmoran[, 5]
mp <- moran.plot(as.vector(scale(map$vble)), nbw)
```

Create a `quadrant` variable for H-H, L-L, H-L and L-H

```{r}
map$quadrant <- NA
map[(mp$x >= 0 & mp$wx >= 0) & (map$lmp <= 0.05), "quadrant"]<- 1
map[(mp$x <= 0 & mp$wx <= 0) & (map$lmp <= 0.05), "quadrant"]<- 2
map[(mp$x >= 0 & mp$wx <= 0) & (map$lmp <= 0.05), "quadrant"]<- 3
map[(mp$x <= 0 & mp$wx >= 0) & (map$lmp <= 0.05), "quadrant"]<- 4
map[(map$lmp > 0.05), "quadrant"] <- 5
```

```{r}
tm_shape(map) + tm_fill(col = "quadrant", title = "",
breaks = c(1, 2, 3, 4, 5, 6),
palette =  c("red", "blue", "lightpink", "skyblue2", "white"),
labels = c("High-High", "Low-Low", "High-Low",
           "Low-High", "Non-significant")) +
tm_legend(text.size = 1)  + tm_borders(alpha = 0.5) +
tm_layout(frame = FALSE,  title = "Clusters")  +
tm_layout(legend.outside = TRUE)
```
